---
title: "Interactive Intro to R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)

source("utils.R")


checker <- function(label,
                    user_code,
                    solution_code,
                    check_code,
                    envir_result,
                    evaluate_result,
                    envir_prep,
                    last_value,
                    ...) {
  check_function <- eval(parse(text = check_code), envir_prep)
  check_function(last_value, user_code)
}

knitr::opts_chunk$set(exercise.checker = checker)
```

```{r}
#```{r child = 'intro_r_background.Rmd'}
#```
```

## What is this? 

The document you're reading is an introduction to programming and data analysis with the R programming language. It's an *interactive* document, where you'll be able to try out R code for yourself. Here, give it a shot to see what `7 + 11` is:

```{r lessons_all_1, exercise=TRUE}
7 + 11
```

Although we're not going to jump into code just yet, this affords us some interesting possibilities. For example, we can run an *exercise*, and you can attempt to solve it: Try writing an expression that results in an output of 30, using the numbers 40 and 7 at least once. (Try a few incorrect answers as well.)

```{r overview3, exercise=TRUE}
# Write your code here!


```

```{r overview3-hint-1, eval=FALSE}
# one possible solution:
40 - (...)
```

```{r overview3-hint-2, eval=FALSE}
# one possible solution:
40 - (7 + ...)
```


```{r overview3-solution}
# one possible solution:
40 - (7 + 3)
```

```{r overview3-check}
function(student_answer, student_code) {
  if(student_answer == 30 && 
     grepl("40", student_code) &&
     grepl("7", student_code)) {
    return(list(
      correct = TRUE,
      message = "Nice work!"
    ))
  } else {
    return(list(
      correct = FALSE,
      message = "Sorry, try again. Make sure you use the numbers 40 and 7 in your code."
    ))
  }
}
```

Sometimes sections will be hidden until you click "Continue":

### Like this

Like this.

## Making Notes and Printing

There's another interesting feature we can take advantage of. You can run the `note()` function in any code block to add a quick note: 

```{r notetest, exercise=TRUE}
note("You can add notes along with other code.")
7 + 3

note("Neat!")
```

The notes themselves can be composed in [Markdown](https://www.markdownguide.org/cheat-sheet/), if you'd like to add more complex formatting:

```{r notetest2, exercise=TRUE}
note("
# Markdown note

This is a markdown note. You can write lists, for examples:

* one
* two
* three

And **bold** or *italic* text.
")
```

